### 2020-04-01 probabilisticProgrammingMartingales.tex
Comments, questions and corrections.

I thought some more about termination in the case of a language with a score function. 
It isn't even clear what termination means in that case.
None of the papers I've found include both unbounded scoring and a definition of termination.
\lo{Can you explain your concern? I don't see any problem. See, e.g., the operational semantics in \citep{MakOP20b}; see also \citep{MakOP20a}.} 
You said that the assumption of a.s. termination was often used. 
Do you have any examples where it's used in a context with unbounded scoring, as going with whatever definition that uses is probably most useful?
\lo{See \citep[Sec.~4, pp.~8 \& 9]{Wagner20} -- I am not sure if this addresses your question.}

@@

To define the reduction relation, let environments be of the form:
\begin{align*}
  E ::= & \, \cdot \mid E t \mid v E \mid \underline{f}(r_1,\dots ,r_{k-1}, E, s_{k+1}, \dots, s_n) \\ & \mid \tY E \mid \tif{E<0}{s_1}{s_2} \mid \tscore (E)
\end{align*}
then a term reduces if it is formed by substituting a redex in an environment i.e.
\begin{align*}
  E[(\lambda x.s) t] & \to E[s[t/x]] \\
  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
  E[\tY v] & \to E[v (\tY v)] \\
  E[\tif{\underline r < 0}{s_1}{s_2}] & \to E[s_1] \text{ where }r < 0 \\
  E[\tif{\underline r < 0}{s_1}{s_2}] & \to E[s_2] \text{ where }r \geq 0 \\
  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1] \\
  E[\tscore(\underline r)] & \to E[\underline r]
\end{align*}

\subsection{Sampling semantics}
This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$. To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples. Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the Borel $\sigma$-algebra and the measure given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element. 
The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.

@@

The limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a normal form. A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined, and it terminates almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.

@@

Having thought a bit more about this, I realise it may not be quite right. I'll see if I can fix it later.}

@@

\section{Ranking functions}
Given a probabilistic program (i.e.~a term $M$), in order to construct a supermartingale to prove its a.s.~termination, a function to assign values to each reachable program state is necessary. Define a ranking function on $M$ to be a measurable function $f:Rch(M) \to \mathbb{R}$ such that
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$
    \item $f(E[\tY N]) \geq 1 + f(E[N(\tY N)])$
    \item $f(E[\tsample]) \geq \int_0^1 f(E[\underline{x}]) \, \mathrm{d}x$
    \item $f(E[R]) \geq f(E[N])$ for any other redex $R$, where $R \to N$.
\end{itemize}

@@

@@

Overall, this implies that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[M_{y_n} \text{not n.f.}]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[M_{y_n} \text{not n.f.}]$ must tend to 0 as $n \to \infty$. $M_{y_n} \text{n.f.} \Rightarrow M_{y_{n+1}} \text{n.f.}$ therefore $\mathbb P[M_{y_n} \text{not n.f.~for infinitely many values of n}] = \mathbb P[\forall n: M_{y_n} \text{not n.f.}] \leq \inf_n \mathbb P[M_{y_n} \text{not n.f.}] = 0$, therefore $M$ terminates almost surely.
\end{proof}

>>>>

I thought some more about termination in the case of a language with a score function. 
It isn't even clear what termination means in that case.
None of the papers I've found include both unbounded scoring and a definition of termination.
\lo{Can you explain your concern? I don't see any problem. See, e.g., the operational semantics in \citep{MakOP20b}; see also \citep{MakOP20a}.} 
You said that the assumption of a.s.~termination was often used. 
Do you have any examples where it's used in a context with unbounded scoring, as going with whatever definition that uses is probably most useful?
\lo{See \citep[Sec.~4, pp.~8 \& 9]{Wagner20} -- I am not sure if this addresses your question.}

@@

\lo{What are the primitive functions $f$? All measurable functions?}

To define the reduction relation, let environments \lo{They are called \emph{evaluation contexts}. Environments mean something else.} be of the form:
\begin{align*}
  E ::= & \, \cdot \mid E t \mid v E \mid \underline{f}(r_1,\dots ,r_{k-1}, E, s_{k+1}, \dots, s_n) \\ & \mid \tY E \mid \tif{E<0}{s_1}{s_2} \mid \tscore (E)
\end{align*}
then a term reduces if it is formed by substituting a redex in an environment i.e.
\begin{align*}
  E[(\lambda x.s) t] & \to E[s[t/x]] \\
  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
  E[\tY v] & \to E[v (\tY v)] \\
  E[\tif{\underline r < 0}{s_1}{s_2}] & \to E[s_1] \text{ where }r < 0 \\
  E[\tif{\underline r < 0}{s_1}{s_2}] & \to E[s_2] \text{ where }r \geq 0 \\
  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1] \\
  E[\tscore(\underline r)] & \to E[\underline r]
\end{align*}

\lo{- I assume this is a CBV language (judging from the (evaluation context) clause $v \, E$). So the first rule should be
\[E[(\lambda x.s) v] \to E[s[v/x]]\]

- In a CBV language, the $\tY$-combinator should only be applied to terms of type $(A \to B) \to (A \to B)$. 
Your rule $E[\tY v] \to E[v (\tY v)]$ will diverge ($E[v (v (v \cdots ))]$).
The redex rule for the Y-combinator should be
\[
\tY (\lambda y.s) \to \lambda z.s[\tY{(\lambda y.s)}/y]\,z \quad \hbox{($z$ is fresh)}
\]

}
\subsection{Sampling semantics}
This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$. To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples. Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the Borel $\sigma$-algebra and the measure, \changed[lo]{denoted $\mu$}, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element. 
The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.
\lo{What does a measurable subset of $S$ look like? Give a ``concrete'' description.}
\lo{Clarify the measure $\mu$: give a ``formula'' for computing $\mu(A)$, for $A \in \Sigma_S$. Is $\mu$ a probability measure?}

@@

The limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a normal form. \lo{Define $\red^n(M,s)$.} 
A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined, and it terminates almost surely if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.
\lo{Why is $\{s \mid M \text{ terminates at } s\}$ a measurable subset of $S$?}
\lo{Why is your definition of a.s.~termination reasonable? 
For example, is it the case that for all $M$, $\mu(\{s \mid M \text{ terminates at } s\}) \leq 1$?}

\lo{Important: Give examples (and non-examples) of a.s.~terminating terms (such as the one you wrote on my whiteboard).}

@@

Having thought a bit more about this, I realise it may not be quite right. I'll see if I can fix it later.}
\lo{I'll read this after you fix it ;-)}

@@

\section{Ranking functions}
Given a probabilistic program (i.e.~a term $M$), in order to construct a supermartingale to prove its a.s.~termination, a function to assign values to each reachable program state is necessary. Define a \emph{ranking function on $M$} to be a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$
    \item $f(E[\tY N]) \geq 1 + f(E[N(\tY N)])$ \lo{Modify this in view of the revised \tY-redex rule.}
    \item $f(E[\tsample]) \geq \int_0^1 f(E[\underline{x}]) \, \mathrm{d}x$

    \lo{I assume you mean $\int_{I} \mathsf{Leb} (\mathrm{d} x) \, f(E[\underline{x}])$.}

    \lo{For the integral to be well-defined, we need $x \mapsto E[\underline x]$ to be a measurable function $\mathbb{R} \to \mathit{Rch}(M)$.}

    \lo{Define the measurable space $\mathit{Rch}(M)$, and the measurable space of SPCF terms.}

    \item $f(E[R]) \geq f(E[N])$ for any other redex $R$, where $R \to N$.
\end{itemize}

@@

\lo{Ongoing}

@@

Overall, this implies that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[M_{y_n} \text{not n.f.}]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[M_{y_n} \text{not n.f.}]$ must tend to 0 as $n \to \infty$. $M_{y_n} \text{n.f.} \Rightarrow M_{y_{n+1}} \text{n.f.}$ therefore 
\begin{align*}
\mathbb P[M_{y_n} \text{not n.f.~for infinitely many values of n}]  
& = \mathbb P[\forall n: M_{y_n} \text{not n.f.}] \\
& \leq \inf_n \mathbb P[M_{y_n} \text{not n.f.}] \\
& = 0
\end{align*} 
therefore $M$ terminates almost surely.
\end{proof}
